Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-12-01 SIMPaper/SIMPaper.tex
added analysis

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjusting for the bias
introduced when sampling a fraction of the study base \(B\) by adding
the offset term \(\log\left(\frac{b}{B} \right)\) as in
\eqref{eqn:offset}. Next, we propose using neural networks to model the
odds.

@@

After case-base sampling, we pass all features, including time, into any
user-defined feed-forward component, to which an offset term is added,
then passed through a sigmoid activation function (Figure
\ref{fig:NNarch}). We use the sigmoid function as its inverse is the
odds, which we can use to calculate the hazard. The general form for the
neural network using CBNN is:

@@

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/cbnnManuscript/figures/nnarch2}

@@

We use two metrics to assess the performance of the different methods of
interest: 1) (IPA) \citep{kattan2018index} and 2) inverse probability
censoring weights-adjusted concordance index (\(C_{IPCW}\))
\citep{uno2011}, which we define below. Both these time-dependent
metrics provide transparency as to when in follow-up time each model may
perform better than the others.

@@

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/cbnnManuscript/analyses/figures/megaPlot}

@@

\caption{Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row demonstrates the $C_{IPCW}$ score for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals).}\label{fig:megaPlot}
\end{figure}

@@

\begin{center}\includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/cbnnManuscript/analyses/figures/megaTable} \end{center}

@@

>>>>

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015}.}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align}

To provide some intuition for this equation, the hazard is the rate at
which the event (\(Y_{i}=1\)) occurs at a given time, conditional on
your survival up to the follow-up-time of interest. As we are looking at
person-moments, we may retrieve the rate at which an event does not
occur (\(Y_{i}=0)\) at a given time by taking the number of samples in
\(b\) and dividing it by the total number of moments in the study base
\(B\) at the current follow-up-time. Though an intuitive approach is
used to describe this equation, a complete derivation of this equation
can be found at \citep{saarela2015}.

The odds of a person-moment being a part of the \emph{case series} is
the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform rate
\(b/B\). Sampling techniques are known to result in a loss of
information. To address this concern, the original case-base methodology
paper for logistic regression uses a measure for relative information
when comparing two averages, based on the size of our base-series and
case-series. If we let the size of \(b=100c\), then the variances and
covariances are expected to be proportional to
\(\frac{1}{c}+\frac{1}{100c}\) rather than
\(\frac{1}{c}+\frac{1}{\inf}\), about 1 percent larger. Concerns were
raised about dependency due to multiple person-moments coming from each
individual, the information loss was studied in Saarela 2015, where they
found a ratio of 100:1, b:c, give an efficiency comparable to
Mantel-Haenszel and conditional logistic regression methods (Saarela
2015). Using \eqref{eqn:main}, we can see how the log-hazard function
can be estimated from the log-odds arising from case-base sampling:
\begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

Sampling techniques are known to result in a loss of information. To
address this concern, the original case-base methodology paper uses a
measure for relative information when comparing two averages, based on
the size of our base-series and case-series. If we let the size of
\(b=100c\), then the variances and covariances are expected to be
proportional to \(\frac{1}{c}+\frac{1}{100c}\) rather than
\(\frac{1}{c}+\frac{1}{\inf}\), about 1 percent larger. If there is a
concern with a 1 percent inflation of these estimates, a larger ratio
than 100 may be used. As there are concerns about dependency due
multiple person-moments coming from each individual, the information
loss was studied in Saarela 2015 (\citep{saarela2015}), where they found
a ratio of 100:1 base-series to case series gives an efficiency
comparable to Mantel-Haenszel and conditional logistic regression
methods \citep{saarela2015}.

To estimate the correct hazard function, we adjusting for the bias
introduced when sampling a fraction of the study base \(B\) by adding an
offset term \(\log\left(\frac{B}{b} \right)\) \eqref{eqn:offset}. Next,
we propose using neural networks to model the odds.

@@

After case-base sampling, we pass all features, including time, into any
user-defined feed-forward component, to which an offset term is added,
then passed through a sigmoid activation function (Figure
\ref{fig:NNarch}). We use the sigmoid activation function as it is the
inverse of the log-odds and commonly available in popular neural network
packages (tensorflow, keras, pytorch, etc) {[}REF?{]}. which we can use
to calculate the hazard. The general form for the neural network using
CBNN is:

@@

{\centering \includegraphics[width=1\linewidth]{../figures/nnarch2}

@@

We use two metrics to assess the performance of the different methods of
interest: 1) (IPA) \citep{kattan2018index} and 2) inverse probability
censoring weights-adjusted concordance index (\(C_{IPCW}\))
\citep{uno2011}, which we define below. We may be interested in an
absolute risk a specific number of years in the future. For the sake of
comparison, we plot each metric over time \(t\) so that the performance
is transparent. As we go further along follow-up-time, we expect the
performance of all models to drop as there are fewer individuals left in
the study by then. For \(C_{IPCW}\), the score demonstrates performance
up to time \(t\) \citep{uno2011}, while IPA is performance at time
\(t\).

@@

{\centering \includegraphics[width=1\linewidth]{../analyses/figures/megaPlot}

@@

\caption{Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row demonstrates the $C_{IPCW}$ score for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals). Note than Cox and CBLR are superimposed.}\label{fig:megaPlot}
\end{figure}

@@

\begin{center}\includegraphics[width=1\linewidth]{../analyses/figures/megaTable} \end{center}

@@

The way neural networks are used in this paper are not directly
interpretable. This is by design, as we only wish to compare to
competitors based on predictive ability. All feed-forward neural network
models can make use of different architectures to improve performance or
interpretability. While a deep neural network as used in this paper is
not directly interpretable, techniques such as Local Interpretable
Model-Agnostic Explanations (LIME) \citep{ribeiro2016model} may be used
to interpret the relevance of each feature to a given prediction. An
example of architectural interpretability is a convolutional neural
network for images, which provide positional information about the
relevant pixels in an image \citep{keras}. Note that neither of these
are specific to CBNN. As CBNN is concerned with learning time-varying
interactions, a concern may be interpreting which time-varying
interactions have been learned. Other secondary methods after the
fitting process may be used to retrieve said information
\url{https://ui.adsabs.harvard.edu/abs/2017arXiv170504977T/abstract}.
